Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-10-26 inference.tex
edits

states, respectively. Constant sets can be written using the set
literal expressions. For e.g., the set $\{1,2\}$ can be written as

@@

semantics-preserving translation to a decidable fragment of
first-order logic.

@@

rely relation $\R$, and the high-level invariant $I$ for the reasons

@@

the same set of records before and after the interference. The
equivalence in $\SL$ translates to equality in first-order logic,
as we describe below. In the inference rules, stability is
enforced constructively by a meta function $\stabilize{\cdot}$, which
accepts an effect and returns a new effect that is guaranteed to be
stable under $\R$.  $\stabilize{\cdot}$ achieves the stability
guarantee by abstracting away the bound $\stg$ in an unstable $\F$ to
an existentially bound $\stg'$ as described below:

@@

Observe that when $\F$ is not stable, then $\stabilize{\F}$ returns a

@@

$I(\stg')$ because all verified transactions preserve the invariant,
and hence only valid database states will ever be witnessed. From the
perspective of RG reasoning, $\stabilize{\cdot}$ effectively weakens
the post-condition of a statement, as done by the

@@

is nonetheless useful because of the $I(\stg')$ assumption on the
existentially bound $\stg'$. The example in

@@

The state transformer rules, like the earlier RG rules, closely follow
the corresponding reduction rules in Fig.~\ref{fig:txnimp}, except
that their language of expression is $\SL$. For instance, while the
reduction rule for \C{UPDATE} declaratively specifies the set of
updated records, the state transformer rule uses $\SL$'s bind
operation to \emph{compute} the set. Other SQL rules do likewise. The
rules for \C{LET} binders, conditionals, and sequences compose the
effects inferred for their subcommands. Thus, the effect of a sequence
of commands $c_1;c_2$ is the union of effects $\F_1$ and $\F_2$ of $c_1$
and $c_2$, respectively, except that $\F_2$ is applied to a state
($\stl$) to which $\F_1$ has already been applied, reflecting their
order of reduction. The inference rule for \C{FOREACH} takes advantage
of the $\SL$'s bind operator to lift the effect inferred for the loop
body to the level of the loop. Since records added to $\stl$ in each
iteration of \C{FOREACH} are independent of the previous iteration
(recall local context independence; Sec. ~\ref{sec:opsem}), sequential
composition of the effects of different iterations is same as their
parallel composition. Since the loop body is executed once per each
$z\in x$, the effect of the the loop is the union of effects ($\F$)
for all $z\in x$, all applied to the same state ($\stl$ and $\stg$).
That is, $\F_{loop}(\stl,\stg) = \bigcup_{z\in
x}\F_{body}(\stl,\stg)$. From the definition of the set monad's bind
operator, $\F_{loop}(\stl,\stg) = x \bind (\lambda
z.~F_{body}(\stl,\stg))$, which mirrors the definition of the rule.
%This development mirrors ideas found in type checkers for certain
%dependent type systems~\cite{KJ14}.

@@

The algorithm ($\mssemof{\cdot}{\cdot}$) shown in Fig.~\ref{fig:logic}
translates an $\SL$ expression ($s$) to FOL. The translation is based
on encoding a set of element type $T$ as a unary predicate on $T$.
The predicate is represented as a meta function that accepts an $x:T$
and returns a quantifier-free proposition that evaluates to true
($\top$) if and only if $x$ is present in the set. Alternatively,
translation may also encode the set as a predicate in the logic
itself, in which case a quantified proposition constraining the
predicate is also generated. For instance, consider the set $\{1,2\}$.
The predicate describing the set can be encoded as the function
$\lambda \v. \v=1 \vee \v=2$, with no further constraints, or it can
be encoded as the function $\lambda \v. g(\v)$ with an associated
constraint ($\phi\in\Prop^1$) defining the uninterpreted boolean
function (i.e., uninterpreted predicate) $g$:
\begin{smathpar}
  \begin{array}{lcl}
    \phi & = & \forall \nu.~g(\nu) \Leftrightarrow \nu=1 \vee \nu=2
  \end{array}
\end{smathpar}
The translation adopts one or the other approach, depending on the
need. For uniformity, we consider the encoding of a set as pair
($\phi$,$\G$), where $\G$ is a meta function, and $\phi$ is a FOL
formula constraining any uninterpreted predicates used in $G$.

@@

The translation algorithm (Fig.~\ref{fig:logic}) takes free variables
into account. Given a set expression $s\in\SL$, whose (possible) free
variables are $\nubar$ in the order of their introduction (top-most
binder first), $\mssemof{\nubar}{s}$ returns the encoding of $s$ as
$(\phi,\G)$.  The meta function $\G$ is a predicate indexed by the
(possible) free variables of $s$, thus, its arity is $|\nubar| + 1$.
Note that $\nubar$ is only a sequence of variables introduced by the
enclosing binders of $s$, and not all may actually occur free in $s$.
Nonetheless, its predicate $\G$ is always indexed by $|\nubar|$ free
variables for uniformity. The translation encodes database state as an
uninterpreted sort. Considering that the state is actually a set of
records, we define an uninterpreted relation ``$\in$'' to relate records
and states. Thus, a variable set expression $\stg$ denoting a database
state is encoded as the predicate $\lambda (\vbar,r).~r\in\stg$, where
$|\vbar|=|\nubar|$ (predicates are uncurried for simplicity; $\vbar$
is a comma-separated sequence; $r\not\in\SL$ is a special variable).
The constraints associated with the encoding of a state are trivial
(denoted $\top$). The set literal expression $\{x\,|\, \varphi\}$ is encoded
straightforwardly. The conditional set expression is encoded as an
if-then-else predicate in FOL, where the predicates on true and false
branches are computed from the set subexpressions $s_1$ and $s_2$,
respectively. The conjunction of constraints $\phi_1$ and $\phi_2$,
from $\mssemof{\nubar}{s_1}$ and $\mssemof{\nubar}{s_2}$ (resp.), is
propagated upwards as the constraint of the conditional expression.
Set union expression is encoded similarly.

>>>>

states, respectively. Constant sets can be written using set
literal expressions. For example, the set $\{1,2\}$ can be written as

@@

semantics-preserving translation amenable for automated verification.

@@

rely relation $\R$, and the high-level invariant $I$ for reasons

@@

the same set of records before and after an interference. The
equivalence in $\SL$ translates to equality in first-order logic, as
we describe below. In the inference rules, stability is enforced
constructively by a meta function $\stabilize{\cdot}$, which accepts
an effect and returns a new effect that is guaranteed to be stable
under $\R$.  $\stabilize{\cdot}$ achieves the stability guarantee by
abstracting away the bound global state ($\stg$) in an unstable $\F$
to an existentially bound $\stg'$ as described below:

@@

Observe that when $\F$ is not stable, $\stabilize{\F}$ returns a

@@

$I(\stg')$ because all verified transactions must preserve the
invariant, and hence only valid database states will ever be
witnessed. From the perspective of RG reasoning, $\stabilize{\cdot}$
effectively weakens the post-condition of a statement, as done by the

@@

is nonetheless useful because of the $I(\stg')$ assumption imposed on
the existentially bound $\stg'$. The example in

@@

The state transformer rules, like the earlier RG rules, closely follow
the corresponding reduction rules in Fig.~\ref{fig:txnimp}, except
that their language of expression is $\SL$. For instance, while the
reduction rule for \C{UPDATE} declaratively specifies the set of
updated records, the state transformer rule uses $\SL$'s bind
operation to \emph{compute} the set. Other SQL rules do likewise. The
rules for \C{LET} binders, conditionals, and sequences compose the
effects inferred for their subcommands. Thus, the effect of a sequence
of commands $c_1;c_2$ is the union of effects $\F_1$ and $\F_2$ of
$c_1$ and $c_2$, respectively, except that $\F_2$ is applied to a
state ($\stl$) to which $\F_1$ has already been applied, reflecting
their order of reduction. The inference rule for \C{FOREACH} takes
advantage of the $\SL$'s bind operator to lift the effect inferred for
the loop body to the level of the loop. Since records added to $\stl$
in each iteration of \C{FOREACH} are independent of the previous
iteration (recall that we make a local context independence assumption
about database programs; Sec. ~\ref{sec:opsem}), sequential
composition of the effects of different iterations is the same as their
parallel composition. Since the loop body is executed once per each
$z\in x$, the effect of the the loop is a union of effects ($\F$)
for all $z\in x$, all applied to the same state ($\stl$ and $\stg$).
That is, $\F_{loop}(\stl,\stg) = \bigcup_{z\in
  x}\F_{body}(\stl,\stg)$. From the definition of the set monad's bind
operator, $\F_{loop}(\stl,\stg) = x \bind (\lambda
z.~F_{body}(\stl,\stg))$, which mirrors the definition of the rule.
%This development mirrors ideas found in type checkers for certain
%dependent type systems~\cite{KJ14}.

@@

The algorithm ($\mssemof{\cdot}{\cdot}$) shown in Fig.~\ref{fig:logic}
translates an $\SL$ expression ($s$) to FOL. The translation is based
on encoding a set of element type $T$ as a unary predicate on $T$.
The predicate is represented as a meta function that accepts an $x:T$
and returns a quantifier-free proposition that evaluates to true
($\top$) if and only if $x$ is present in the set. Alternatively,
the translation may also encode the set as a predicate in the logic
itself, in which case a quantified proposition constraining the
predicate is also generated. For instance, consider the set $\{1,2\}$.
The predicate describing the set can be encoded as the function
$\lambda \v. \v=1 \vee \v=2$, with no further constraints, or it can
be encoded as the function $\lambda \v. g(\v)$ with an associated
constraint ($\phi\in\Prop^1$) defining the uninterpreted boolean
function (i.e., uninterpreted predicate) $g$:
\begin{smathpar}
  \begin{array}{lcl}
    \phi & = & \forall \nu.~g(\nu) \Leftrightarrow \nu=1 \vee \nu=2
  \end{array}
\end{smathpar}
The translation adopts one or the other approach, depending on the
need. For uniformity, we consider the encoding of a set as pair
($\phi$,$\G$), where $\G$ is a meta function, and $\phi$ is a FOL
formula constraining any uninterpreted predicates used in $G$.

@@

The translation algorithm (Fig.~\ref{fig:logic}) takes free variables
into account. Given a set expression $s\in\SL$, whose (possible) free
variables are $\nubar$ in the order of their introduction (top-most
binder first), $\mssemof{\nubar}{s}$ returns the encoding of $s$ as
$(\phi,\G)$.  The meta function $\G$ is a predicate indexed by the
(possible) free variables of $s$, and thus its arity is $|\nubar| + 1$.
Note that $\nubar$ is only a sequence of variables introduced by the
enclosing binders of $s$, and not all may actually occur free in $s$.
Nonetheless, its predicate $\G$ is always indexed by $|\nubar|$ free
variables for uniformity. The translation encodes database state as an
uninterpreted sort. Considering that the state is actually a set of
records, we define an uninterpreted relation ``$\in$'' to relate records
and states. Thus, a variable set expression $\stg$ denoting a database
state is encoded as the predicate $\lambda (\vbar,r).~r\in\stg$, where
$|\vbar|=|\nubar|$ (predicates are uncurried for simplicity; $\vbar$
is a comma-separated sequence; $r\not\in\SL$ is a special variable).
The constraints associated with the encoding of a state are trivial
(denoted $\top$). The set literal expression $\{x\,|\, \varphi\}$ is encoded
straightforwardly. The conditional set expression is encoded as an
if-then-else predicate in FOL, where the predicates on true and false
branches are computed from the set subexpressions $s_1$ and $s_2$,
respectively. The conjunction of constraints $\phi_1$ and $\phi_2$,
from $\mssemof{\nubar}{s_1}$ and $\mssemof{\nubar}{s_2}$ (resp.), is
propagated upwards as the constraint of the conditional expression.
A set union expression is encoded similarly.
